\documentclass{article}
\setcounter{errorcontextlines}{10000}
\usepackage{fontspec}
\usepackage{xunicode}
\usepackage{xltxtra}

\usepackage{fullpage}
\usepackage{mleftright}

\usepackage{hyperref}
\hypersetup{%
	linktocpage  = true, %page number is the link (not title)
	colorlinks   = true, %Colours links instead of ugly boxes
	urlcolor     = darkblue, %Colour for external hyperlinks
	linkcolor    = darkblue, %Colour of internal links
	citecolor    = darkblue   %Colour of citations
}

% define colors %
\usepackage[table]{xcolor}

\definecolor{lightgray}{rgb}{0.9,0.9,0.9}
\definecolor{darkgray}{rgb}{0.4,0.4,0.4}
\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
\definecolor{darkblue}{rgb}{0.02, 0.17, 0.40}

% good looking urls
\urlstyle{same}

\newcommand{\theoremname}{Theorem}
\newcommand{\corollaryname}{Corollary}
\newcommand{\problemname}{Problem}
\newcommand{\conjecturename}{Conjecture}
\newcommand{\definitionname}{Definition}

\usepackage{url}
\usepackage{amsmath,amsthm,amsfonts,mathtools}
\usepackage{xfrac}

\newtheorem{theorem}{\theoremname}
\newtheorem{corollary}{\corollaryname}
\newtheorem{problem}{\problemname}
\newtheorem{conjecture}{\conjecturename}
\newtheorem{definition}{\definitionname}

\title{Computability and Complexity:\\Homework December 2017}
\author{Aurélien Ooms\footnote{aureooms@ulb.ac.be}}
\date{\today}

\begin{document}
\maketitle
\section*{Instructions} The homework is due on Monday December 18 10AM at the
beginning of the Q\&A session. Your solution must be typeset and printed on
paper. A correct solution will grant you bonus points for the final exam.
Plagiarism will be severely sanctionned.

\section*{Homework}
Let us start with a definition:
\begin{definition}
  Let $\phi$ be a 3CNF-formula. A $\ne$-assignment to the variables of $\phi$
  is one where each clause contains two literals with unequal truth values.
\end{definition}
In other words, a $\ne$-assignment satisfies $\phi$ without assigning three
true literals in any clause.

Let $\ne$SAT be the collection of 3CNF-formulas that have a $\ne$-assignment.
\begin{definition}
  $\ne$SAT = $\{\,\langle \phi \rangle\colon\, \text{$\phi$ has a $\ne$-assignment.} \,\}$
\end{definition}

\paragraph{Your task:} You are asked to prove that $\ne$SAT is NP-complete.

\paragraph{Hint:} Prove that the negation of any $\ne$-assignment of $\phi$ is
also a $\ne$-assignment of $\phi$.
%
\end{document}
