\documentclass{article}
\errorcontextlines 10000

\makeatletter

\usepackage{fontspec}
\usepackage{xunicode}
\usepackage{xltxtra}

\usepackage{fullpage}
\usepackage{mleftright}

\usepackage{hyperref}

\hypersetup{%
	linktocpage  = true, %page number is the link (not title)
	colorlinks   = true, %Colours links instead of ugly boxes
	urlcolor     = darkblue, %Colour for external hyperlinks
	linkcolor    = darkblue, %Colour of internal links
	citecolor    = darkblue   %Colour of citations
}

%\hypersetup{
%	linktocpage  = true, %page number is the link (not title)
%	colorlinks   = true, %Colours links instead of ugly boxes
%	allcolors = bookColor,
%	hidelinks = true
%}

% good looking urls
\urlstyle{same}

\newcommand{\theoremname}{Theorem}
\newcommand{\corollaryname}{Corollary}
\newcommand{\problemname}{Problem}
\newcommand{\conjecturename}{Conjecture}
\newcommand{\definitionname}{Definition}

% we use this for our references as well
\let\sref\ref
\AtBeginDocument{\renewcommand{\ref}[1]{\mbox{\autoref{#1}}}}
\usepackage[nameinlink]{cleveref}

% define colors %
\usepackage[table]{xcolor}

%\usepackage[table,dvipdfx,cmyk]{xcolor}
%\definecolor{bookColor}{cmyk}{0 ,0 ,0 ,1}
%\color{bookColor}

\definecolor{lightgray}{rgb}{0.9,0.9,0.9}
\definecolor{darkgray}{rgb}{0.4,0.4,0.4}
\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
\definecolor{darkblue}{rgb}{0.02, 0.17, 0.40}
\usepackage[stable]{footmisc}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}
\usepackage{amsmath,amsthm,amsfonts,mathtools}
\usepackage{xfrac}
\newtheorem{theorem}{\theoremname}
\newtheorem{corollary}{\corollaryname}
\newtheorem{problem}{\problemname}
\newtheorem{conjecture}{\conjecturename}
\newtheorem{definition}{\definitionname}

% nicely spaced operators
\DeclareMathOperator{\Exists}{\exists}
\DeclareMathOperator{\Forall}{\forall}

\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\newcommand{\card}[1]{|#1|}

\newcommand{\st}{\colon\,}
\newcommand{\TM}{TM}
\newcommand{\Atm}{A\textsubscript{TM}}
\newcommand{\HALTtm}{HALT\textsubscript{TM}}

\newcommand\N{\mathbb{N}_1}

\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}



% TURING MACHINE DESCRIPTIONS
\usepackage{changepage}
\newenvironment{steps}%
{%
\vspace{0.25cm}%
\begin{adjustwidth}{0.3cm}{0cm}%
\begin{description}%
}
{%
\end{description}%
\end{adjustwidth}%
\vspace{0.1cm}%
}

\newcounter{TMachine}[section]
\renewcommand{\theTMachine}{\thesection.\arabic{TMachine}}%
\newenvironment{TMachine}[1]
  {\refstepcounter{TMachine}%
   \par%
   \vspace{.5\baselineskip\@plus.2\baselineskip\@minus.2\baselineskip}% Space above
   \noindent{#1}%
   \begin{steps}}%\begin{TMachine}
  {\end{steps}%
\vspace{.5\baselineskip\@plus.2\baselineskip\@minus.2\baselineskip}% Space below
}% \end{TMachine}

\newcommand{\accept}{\emph{accept}}
\newcommand{\reject}{\emph{reject}}

\makeatother

\title{Computability and Complexity:\\Exercise Session 5 (2015-10-23)}
\author{Aurélien Ooms\footnote{aureooms@ulb.ac.be}}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{3SAT is polynomial-time reducible to CLIQUE\footnote{%
Theorem 7.32 from the reference book: Sipser M.,
\emph{Introduction to the Theory of Computation}, 3rd edition (2013).}}

\subsection{Introduction}

Let us recall the big picture. We have shown in the first part of the course,
the part about computability, that not all problems can be solved by means of
Turing machines. We made the distinction between decidable and undecidable
problems, the former being the ones we can solve with algorithms, that is,
procedures that eventually terminate.

The second part, that is, the part about complexity, focuses on the resources
it takes to solve decidable problems. We already introduced the classes P and
NP. NP is the class of problems for which there exists a nondeterministic
polynomial-time algorithm. An equivalent definition is to consider the
languages for which all words have at least one corresponding certificate of
appartenance of polynomial size that can be verified using a polynomial-time
algorithm called a verifier. P is the class of problems for which there exists
a deterministic polynomial-time algorithm. By definition P is contained in NP.
However, we do not know whether P is different from NP.

Later on in the lectures, we will define the notion of NP-completeness and show
that, while we (today) cannot answer the question ``P vs. NP'', we can identify
a set of problems, the NP-complete problems, that capture the complexity of all
the problems in NP. This kind of result relies heavily on the notion of
reduction.

\subsection{Definitions}

We already saw that the reducibility of a problem to
another need not rely on similarity between the two problems. It is
possible to prove reducibility results between two apriori unrelated problems.
The goal of this exercise is to show you a first reduction of an NP-complete
problem to a problem in NP that, at first sight, might not seem intuitive.

Let us begin with a few definitions
\begin{definition}
	A boolean formula \(\phi\) over \(n\) variables \(x_1,x_2,\ldots,x_n\) is in
	conjunctive normal form if \(\phi\) can be
	written as a conjunction of \(m\) clauses \(c_1 \land c_2 \land \cdots \land
	c_m\) where each clause \(c_j\) can be written as a disjunction of
	\(k_j\) literals
	\((l_{j1} \lor l_{j2} \lor \cdots \lor l_{jk_j})\) where each literal is
	either a variable \(x_i\) or the negation of a variable \(\lnot x_i\).
	The language CNF is the language of boolean formulae in conjunctive normal
	form.
\end{definition}

The language CNF-SAT is defined as
\begin{definition}
	CNF-SAT \(= \{ \langle \phi \rangle \st \phi \in \text{CNF and there
						exists an assignment } \in
						\{0,1\}^n \text{ on the
							variables } x_1,x_2,\ldots,x_n \text{ that makes the formula evaluate to
		true.}\}\)
\end{definition}

The language 3SAT is a subset of CNF-SAT
\begin{definition}
	3SAT \(= \{ \langle \phi \rangle \st \phi \in \text{CNF-SAT and has
	exactly 3 literals per clause.}\}\)
\end{definition}

The language CLIQUE is the following
\begin{definition}
	CLIQUE \(= \{ \langle G , k \rangle \st \text{ \(G\) is a graph with a clique
	of size \(k\).}\}\)
\end{definition}

\subsection{Statement and proof}

We now prove the following
\begin{theorem}
	3SAT \(\le_P\) CLIQUE
\end{theorem}

\begin{TMachine}{\(f\) : on input \(\langle \phi \rangle\):}
\item[1.] \(V \gets \emptyset\)
\item[2.] \(E \gets \emptyset\)
\item[3.] For each \(c_j = (l_{j1},l_{j2},l_{j3}) \in \phi\):
\item[3.1.] \(T \gets \{v_{j1},v_{j2},v_{j3}\}\)
\item[3.2.] \(V \gets V \cup T\)
\item[3.3.] For each \(v_{jk} \in T\):
\item[3.3.1.] For each \(v_{j'k'} \in V \setminus T\):
\item[3.3.1.1.] \(E \gets E \cup \{\{v_{jk},v_{j'k'}\}\}\) if \(l_{jk} \neq \lnot l_{j'k'}\)
\item[4.] Output \(\langle (V,E) , m \rangle\)
\end{TMachine}

\subsubsection{Complexity}
An important step when reducing an NP-complete problem to an NP problem is to
check that the reduction takes polynomial time and that the new problem is of
polynomial size.

It is not difficult to see that the complexity of the algorithm above is
quadratic in the number of clauses (\(O(m^2)\)). The size of the output graph
(\(\card{V} + \card{E}\)) is also \(O(m^2)\). \(f\) runs thus in polynomial
time and produces a graph of polynomial size.

\subsubsection{Correctness}

We need to prove that \(\phi \in \text{3SAT} \iff f(\langle \phi \rangle) \in
\text{CLIQUE}\). We prove the two implications of the equivalence.

\paragraph{(\(\implies\))} If \(\langle\phi\rangle \in \text{3SAT}\) then, by definition,
there exists an assignment \((x_1', x_2', \ldots, x_n') \in \{0,1\}^n\) such
that for each clause there is at least one literal \(l_{jz_j}\) that evaluates
to true. In other words, \(\exists l_{1z_1},l_{2z_2},\ldots,l_{mz_m}\) such
that no two literals \(l_{az_a},l_{bz_b}\) are contradictory, that is,
\(l_{az_a} \neq \lnot l_{bz_b} \Forall a,b \in [m]\). The corresponding
vertices in the graph of \(f(\langle \phi \rangle)\) form a \(m\)-clique, that
is, the graph induced by \(v_{1z_1},v_{2z_2},\ldots,v_{mz_m}\) must be
complete. This holds because, by definition, an edge can only be missing if two
vertices represent literals of the same clause, which is not the case here
since we have selected exactly one literal per clause, \textbf{or} if two
vertices represent contradictory literals, which by construction is not
possible either. Hence, \(f(\langle \phi \rangle) \in \text{CLIQUE}\).

\paragraph{(\(\impliedby\))} If \(f(\langle \phi \rangle) = \langle G, m \rangle \in \text{CLIQUE}\),
then, by definition, \(G\) must contain a clique of size \(m\). By
construction of \(G\), this clique cannot contain more than one vertex per
triple, since there is no edge between two vertices of the same triple. Since
there are \(m\) triples, a clique of size \(m\) must thus contain exactly one
vertex per triple. Let us name those vertices
\(v_{1z_1},v_{2z_2},\ldots,v_{mz_m}\), they correspond to literals
\(l_{1z_1},l_{2z_2},\ldots,l_{mz_m}\) covering all the clauses of \(\phi\).
By construction of \(G\), two such literals cannot be contradictory because
their corresponding vertices, being part of the same clique, have an edge
between them. Hence, \(\langle \phi \rangle \in\) 3SAT.

\subsection{About Karp reductions}

What we just did is called a Karp reduction. Given two problems \(A\)
and \(B\), an algorithm for problem \(B\), and a polynomial-time
mapping procedure \(f\) to translate any
instance \(a\) of \(A\) into an instance \(b = f(a)\) of \(B\) in such a way that the
answer to the decision problem \(a\) is ``yes'' if and only if the answer to
the decision problem \(b\) is ``yes'', we can decide any instance of \(A\) using
the following algorithm
\begin{TMachine}{On input \(\langle a \rangle\):}
\item[1.] Translate instance \(a \in A\) to the corresponding instance \(b \in
	B\) in polynomial time.
\item[2.] Answer the same as the algorithm for \(B\) run on \(\langle b \rangle\).
\end{TMachine}
Moreover, if the algorithm for \(B\) runs in polynomial time so does this
algorithm.

\subsection{CLIQUE is NP-complete}

During the next lecture, we will define the notion of NP-completeness
\begin{definition}
	A language \(B\) is NP-complete if it satisfies two conditions:
	\begin{enumerate}
		\item \(B\) is in NP, and
		\item every \(A\) in NP is polynomial-time reducible to \(B\).
	\end{enumerate}
\end{definition}
and see that
\begin{theorem}
	CNF-SAT is NP-complete
\end{theorem}
and that
\begin{theorem}
	CNF-SAT \(\le_P\) 3SAT.
\end{theorem}
Those two theorems together with the one we just proved and the fact that
CLIQUE is in NP\footnote{Left as exercise for the reader}
imply the following corollary
\begin{corollary}
	CLIQUE is NP-complete.
\end{corollary}
\end{document}
